---
title: "Chemical Sample Classification Report"
subtitle: "Findings"
author: "Petter Lövehagen"
date: "09 May 2023"

RVersion: 3.6.0 
RStudio: 2023.03.0+386 Cherry Blossom Release
Platform: "x86_64-w64-mingw32/x64 (64-bit)"
OS: "Windows 11 SE"

format: 
  docx: default
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-line-numbers: true
    theme: spacelab
    html-math-method: mathjax
    highlight: "tango"
    #css: "../data/custom.css"
prefer-docx: true
lang: "en-GB"
subject: "chemical sample classifition"
#bibliography: "../data/AS_CW_References.bib"
csl: "../data/harvard-university-of-the-west-of-england.csl"

knitr:
  opts_chunk:
    include: false
    tidy: true
    echo: false
    warning: false
    message: false
    error: true
    fig.align: left
    out.width: '90%'
    fig.width: 6
    fig.height: 4
---

```{r final_packages}
# load/install R packages/libraries
# set seed
# load flextable defaults
# see R file for details
set.seed(567)
file <- "classify"
source("R/chem_report_load.R")
```

## Findings {#sec-findings}

### Data {#sec-data-comment}

The supplied sample dataset is of good quality and generally suitable for classification modelling. There are some outstanding questions about missing data and potential outliers, but nothing of or which prevents analysis and modelling.

### Data Reduction {#sec-data-reduction-find}

Principal Components Analysis was found to be unsuitable as a dimension reduction technique on this data - not enough variability was captured.

Feature Reduction, where variables are excluded from the classification models, was found to be successful at a small cost to performance.

### Classification Models {#sec-classification-model-find}

Trees, random forests, discriminant analysis, support vector machine, clustering and k nearest neighbours were investigated.

The best performing model overall was a Discriminant Analysis model on the using all variables, resulting in an accuracy of 97.7% on the validate set.

The best performing model using a reduced variable dataset was also a DA model, with an accuracy of 97.2%.

```{r load_model_data}
# load scaling function
source("R/scale_data_using_train_stats.R")


#load test data
test <- read.csv("../data/chem_test.csv")
num_test <- test[,-21]
#dim(test)


# calculate train means and sds for scaling function
train_label <- read.csv("../data/chem_train.csv")
train <- train_label[,-21]
train_means <- colMeans(train)
train_sds <- apply(train, 2, sd)

# scale based on train stats
test_scaled <- scale_data_using_train(num_test, train_means = train_means, train_sds = train_sds)
#dim(test_scaled)

test[,21]

```

```{r predict_test_complete}

# load models
DA_complete_model_1 <- readRDS("../report/DA_model_1.RDS")
DA_reduced_model_7 <- readRDS("../report/DA_model_7.RDS")

# make class predictions on new dataset
predictions <- predict(DA_complete_model_1, newdata = test_scaled, type = "class")


# compare predictions to true class
accuracy <- sum(predictions$classification == as.factor(test[,21])) / nrow(test_scaled)

# print accuracy
cat(paste0("Accuracy of Complete_Var model on test dataset: ", round(accuracy, 3), "\n"))

comp_acc <- round((accuracy*100),2)
```

```{r predict_test_reduced}
#"X7"  "X8"  "X9"  "X10" "X3"  "X11" (col order)

# variables
test_reduced_3 <- test_scaled[, c("X7", "X8", "X9", "X10", "X3", "X11")]



# load models
DA_reduced_model_7 <- readRDS("../report/DA_model_7.RDS")

# make class predictions on new dataset
predictions2 <- predict(DA_reduced_model_7, newdata = test_reduced_3, type = "class")

# compare predictions to true class
accuracy2 <- sum(predictions2$classification == as.factor(test[, 21])) / nrow(test_scaled)

# print accuracy
cat(paste0("Accuracy of Reduced_Var model on test dataset: ", round(accuracy2, 3), "\n"))

reduce_acc <- round((accuracy2*100), 2)


```

### Test Data

The final assessment of the models is on completely untouched data - the `test` set.

::: callout-important
## Test Results

-   Accuracy of DA Model using **All Variables**: `r comp_acc`%

-   Accuracy of DA Model using **Reduced Variables**: `r reduce_acc`%

Performance hit from using less variables: `r comp_acc - reduce_acc`%
:::

<br>

<br>

::: callout-important
## Recommendations

-   Exploratory session with domain expert to gain insight into dataset:
    -   Missing data
    -   Outliers
    -   Variables - especially label E
    -   Measurements
-   Deploy Classification Model using reduced variable set
-   Consider further model improvement
    -   Speed gains

    -   Accuracy gain
:::
