---
title: "Chemical Sample Classification Report"
subtitle: "Data Preparation"
author: "Petter Lövehagen"
date: "08 May 2023"

RVersion: 3.6.0 
RStudio: 2023.03.0+386 Cherry Blossom Release
Platform: "x86_64-w64-mingw32/x64 (64-bit)"
OS: "Windows 11 SE"

format: 
  docx: default
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-line-numbers: true
    theme: spacelab
    html-math-method: mathjax
    highlight: "tango"
    #css: "../data/custom.css"
prefer-docx: true
lang: "en-GB"
bibliography: "../data/AS_CW_References.bib"
csl: "../data/harvard-university-of-the-west-of-england.csl"

knitr:
  opts_chunk:
    include: false
    tidy: true
    echo: false
    warning: false
    message: false
    error: true
    fig.align: 'left'
    out.width: '90%'
    fig.width: 6
    fig.height: 4
---

<!--# The `data` qmd file uses the following libraries: dplyr (data manipulation), flextable (tables), naniar (missing data). -->

<!--# The 'chem_packages.R file installs/loads required packages, sets a seed (for reproducibility) and loads flextable defaults. -->

```{r data_packages}
# load/install R packages/libraries
# set seed
# load flextable defaults

file <- "data"
source("chem_packages.R")

# library(dplyr)
# library(flextable)
# library(naniar)

```

```{r flextable_defaults}
# 
# set_flextable_defaults(
#   font.family = "Arial",
#   font.size = 10, 
#   font.color = "#333333",
#   border.color = "gray", 
#   background.color = "#FAFAFA",
#   table.layout= "autofit",
#   theme_fun = theme_vanilla,
#   padding.top = 3, padding.bottom = 3,
#   padding.left = 4, padding.right = 4,
#   na_str = "<na>", 
#   digits = 2
#   )


```

```{r inspect_data}

# load csv file
chem_data <- read.csv("../data/5976423.csv")

# display first rows
head(chem_data)

# check dimensions of df
dim(chem_data)
```

## Data {#sec-data}

The supplied dataset contains `r nrow(chem_data)` measurements of `r ncol(chem_data)-1` numerical variables.

The analysis team and report writer did not have any contextual information about the data, e.g.:

-   what has been measured
-   variable importance, interactions
-   measurement scale, units, reliability

Therefore, variables are treated equally and without bias in the analysis and modelling. Decisions are made on statistical grounds and stated assumptions.

::: callout-tip
## Tip

Additional domain-specific information about the data could result in alternative modelling decisions and outcomes and should be considered by the client.
:::

### Missing data {#sec-missing-data}

```{r missing_check}
# check for missing data

total_na<-sum(is.na(chem_data))
pct_na<-round(mean(is.na(chem_data)) * 100,2)
total_na
pct_na
```

```{r missing_summaries}
# create a new data frame with only rows with missing data
missing_rows <- chem_data[apply(chem_data, 1, function(row) any(is.na(row))), ]

# print the missing rows data frame
#print(missing_rows)

# dimensions of missing rows
#dim(missing_rows)

# summarise missing values per column 
na_counts <- colSums(is.na(missing_rows))
  
# calculate percentage of missing values per variable
na_percent <- round(na_counts/nrow(chem_data)*100, 2)

# summarise missing values by label
na_summary <- chem_data %>%
  group_by(label) %>%
  summarize_all(~ sum(is.na(.)))

total_obs <- nrow(chem_data)

row_sums <- data.frame(Label = na_summary$label, 
                        Count = rowSums(na_summary[,-1]),
                        Percentage = round(rowSums(na_summary[,-1])/total_obs * 100, 2))

#max label
max_label_pct <- row_sums$Label[which.max(row_sums$Percentage)]
max_label_cnt <- row_sums$Label[which.max(row_sums$Count)]



# Flextable for Word docx
missing_by_var <- data.frame(Variable = names(na_counts), Count = na_counts, Percentage = na_percent)

FT_missing_by_var <- flextable(missing_by_var)


FT_top_missing_vars <- missing_by_var |>
  arrange(desc(Count)) |>
  head(5) |>
  flextable() |>
  bold(j=1) 


#FT_top_missing_vars <- align(FT_top_missing_vars, align = "left")


FT_missing_by_label <- flextable(row_sums) |>
  bold(j=1)


FT_missing_table <- flextable(missing_rows)

```

```{r missing_naniar}

# visualise missing pattern
vis_miss(chem_data)

# visualise missing pattern
gg_miss_var(chem_data)

# visualise with facet
facet_plot <- gg_miss_fct(chem_data, fct = label)
facet_plot
# pairwise correlation between missingness
upset_plot <-gg_miss_upset(chem_data)
upset_plot

# amount and percentage missing for each variable
miss_var<-miss_var_summary(chem_data)

# amount and percentage missing for each observation
miss_case<-miss_case_summary(chem_data)

# var with most missing
most_missing_var <- miss_var %>%
  arrange(desc(n_miss)) %>%
  slice(1) %>%
  pull(variable)
most_missing_var


```

Before proceeding with [Exploratory Data Analysis](#sec-exploratory-data-analysis), it is important to explore the extent of missing data and whether there are patterns to the `missingness`. Some classification models assume no missing data, and depending on the amount, prevalence and patterning of missing data, different assumptions and techniques can be applied.

We investigate `missingness` to see whether the assumption that missing data is 'missing at random' (MAR) holds - that is, is the probability of `missingness` only dependent on **observed** variables? If so and there is a significant amount of missing data (\>5%), multiple imputation approaches may be explored to *impute* missing values by estimation through statistical inference.

<!--# Missingness may require additional analysis asa future datasets may not be suited to discarding rows with missing values.  It is assumed that the dataset is 'missing at random although this should be confirmed to ensure that assumptions hold.  -->

<!--# The mcar_test() from the `naniar` package has been run and it leads us to reject the null hypothesis that the data is `missing completely at random`.  Additional tests and explorations could be employed to check`missing at random` -->

```{r mcar}
# missing completely at random
# results (statistic = 416, p < 0.05) lead to rejecting H0 that data is MCAR and accepting alternative that data is not MCAR.
mcar_test(chem_data)


```

`r case_when(pct_na <= 5 ~ paste0("The sample has ",total_na," rows with at least one missing value, which is ",pct_na,"% of the entire dataset. As there are fewer missing values than the 5% threshold, imputation is not required and removing these rows is acceptable."), pct_na > 5 ~ paste0("The sample has ",total_na," rows with at least one missing value, which is ",pct_na,"% of the entire dataset. As there are more missing values than the 5% threshold, how to handle missing data needs further consideration.") )`

```{r vis_miss, echo=FALSE, include=TRUE,message=FALSE, warning=FALSE, fig.align='left'}
#| fig-alt: "Summary Visualisation of Missing Data"
#| label: fig-missData
#| fig-cap: "Visualised Missing Data"


# visualise missing pattern
vis_miss(chem_data)
```

If the missing data is not MAR - that is, it appears to depend on **unobserved** variables, there may be other, more appropriate imputation methods such as maximum likelihood imputation. Unobserved influences in this scenario may include the sample purity/quality, sample handling/measuring differences, measurement/equipment discrepancies or chemical compositions of the samples.

*Imputation* can introduce bias into the analysis if the assumptions of the imputation method are violated, or the imputed values differ significantly from true missing values.

::: callout-caution
The client should consider `missingness` in their data:

-   Are there any concerns regarding missing data?
-   Is the equipment, staff, technique, facilities, chemicals, etc. consistent?
-   What are the reasons for missing data? Can they be mitigated?
-   Can data be collected where it has been identified as missing?
-   Is there likely to be (more, less, similar amount of) missing data in the future?
:::

#### Missing value details by class and variable {#sec-missing-value-details-by-class-and-variable}

```{r var_missing_sentence, echo=FALSE, results='asis'}
# vars with missing data
missingVars <- miss_var_which(chem_data)
# how many variables with missing data
missingLen <- length(missingVars)

# initialise output variable
output <- ""

# ifelse to print different output depending on how many variables with missing data. 
if (missingLen == 0) {
  output <- "This sample has no variables with missing values."
} else if (missingLen == 1) {
  output <- paste0("In this sample, ", missingVars, " is the only variable with missing data.")
} else if (missingLen < 6) {
  # create a character vector of the missing variables, with commas between each item except for the last
  missingVars_string <- paste(missingVars[-length(missingVars)], collapse = ", ")
  # Add "and" to the end of the character vector
  missingVars_string <- paste0(missingVars_string, " and ", missingVars[length(missingVars)])
  # Print the sentence
  output <- paste0("In this sample, the following variables have missing data: ", missingVars_string, ".")
} else {
  output <- paste0("This dataset has ", missingLen, " variables with missing data. The five variables with the most missing data are:")
  
}

# Print the output variable
cat(output)

```

`r output`

```{r print_FT_top_missing_vars, include=TRUE, echo=FALSE, results='asis'}
#| tbl-cap: "Top 5 Variables with missing data, by count"
#| tbl-cap-location: bottom
#| tbl-alt: "Table with top five variables with most missing data"
#| tab.id: tbl-top5missing
if (missingLen >= 6) {
  FT_top_missing_vars
}
```

Variable `r most_missing_var` has the most missing values with `r miss_var[which.max(miss_var$n_miss), "n_miss"]`, accounting for `r max(miss_var$pct_miss)`% of this variable's data.

No single sample (observation) has more than `r max(miss_case$n_miss)` missing values. This means that, at most, a sample is missing `r round(max(miss_case$pct_miss), 2)`% of its data.

The upset plot below shows the five variables with the most missing values and confirms that there are no instances where there are missing values in two variables in the same observations.

```{r upset_plot, include=TRUE, echo=FALSE, message = FALSE, tab.align='left', fig.alt="Upset plot of missing data"}
#| label: fig-upsetplot
#| fig-cap: "Upset plot of missing data"

# call upset plot
upset_plot
```

If we group missing data by the target label, we see that `r max_label_pct` has the most missing data at `r max(row_sums$Percentage)`%.

```{r missingByLabel, echo=FALSE, include=TRUE,message=FALSE, warning=FALSE, results='asis'}
#| tbl-cap: "Missing Data by Class (label)"
#| tbl-cap-location: bottom
#| tbl-alt: "Table with missing data, grouped by chemical class"
#| 
FT_missing_by_label

```

Patterns of `missingness` begin to emerge when examining missing data for variables, grouping by 'labels'. It seems that there is a relationship between missing values for sequential variables when grouping observations. This is clearly visible in the `heatmap` below as well as in the stepped pattern in @fig-missData above.

```{r heatmapMissing, include=TRUE, echo=FALSE, message = FALSE}
#| fig-alt: "Heatmap of missing data per variable, grouped by Chemical class (label)"
#| label: fig-heatmapMiss
#| fig-cap: "Heatmap of missing data per variable, grouped by Chemical class"

facet_plot
```

It is important to note that the target variable ('label') has no missing values.

::: callout-warning
Is there anything systemic, in terms of `missingness`:

-   X1-X4 is *only* missing for label A
-   X6, X8 are *only* missing for label B
-   X9-X12 is *only* missing for label C
-   X13-X16 are *only* missing for label D
-   X17-X19 are *only* missing for label E

Is there a relationship between the variables, in terms of what they measure?
:::

::: callout-important
## Decision

Given the minimal amount of missing data - it will be removed before exploratory data analysis, data splitting and modelling.

If this dataset is representative, the missing data may not be an issue in the future - assuming the pattern identified above is understood.
:::

### Splitting data

```{r cleandata}
clean_chem_data <- chem_data[complete.cases(chem_data), ]
#clean_chem_data
dim(clean_chem_data)
sum(is.na(clean_chem_data))
```

Before proceeding with additional exploratory analysis, we split the dataset into three random subsets:

-   `train` - used to *train* the classification model to *learn* the relationship between the variables and the label
-   `validation` - used to *validate* the performance of the trained model(s) and to tune any hyperparameters; determines how well the model generalises to 'unseen' data
-   `test` - used to *evaluate* the final performance of the model post training and tuning; this data is only used once and kept separate.

In order to not introduce bias into the modelling process, the data will be split prior to any EDA, where the dataset is explored and summarised. Any insights or observations made during EDA will emerge solely from the training subset.

The only thing we check in advance is the target variable balance - that is the split across labels. Given that it is not *perfectly* balanced, the data subsets will be split with 'stratification' ensuring that each partition has a representative proportion of each class.

```{r frequency, include=TRUE, echo=FALSE, message = FALSE, results='asis'}
#| label: tbl-countByClass
#| tbl-cap-location: bottom
#| tbl-cap: "Frequency Table by Class (label)"
#| tbl-alt: "Frequency table grouped by chemical class"

# frequency table
freq<-clean_chem_data %>%
  dplyr::count(label, name = 'Frequency') %>% 
  mutate(Percent = round(Frequency / sum(Frequency[1:5])*100, 2)) 

freq |>
  flextable() |>
  autofit() |>
  bold(j=1)


```

::: callout-important
## Decision

The clean dataset is sufficiently large to split:

-   `train` - 50%
-   `validation` - 25%
-   `test` - 25%
:::

<!--# The below is a function to split a dataframe into three random subsets - train, validate, test.  It takes `dataframe`, `target column` (e.g. label), `proportions` (train, validate, test - need to sum to 1), `stratify` (TRUE, FALSE whether to stratify on target column).  It is saved in a separate file -->

```{r load_df_splitter}
#load dataset splitter function
source("df_splitter.R")

```

```{r applySplitCleanChem}
# apply to clean_chem_data
datasets <- df_splitter(df = clean_chem_data, target_col = "label", train_prop = 0.5, val_prop = 0.25, test_prop = 0.25, stratify = TRUE)

# print the sizes of each resulting dataset
cat("train:", nrow(datasets$train), "\n")
cat("validation:", nrow(datasets$val), "\n")
cat("test:", nrow(datasets$test), "\n\n\n")

# check that the proportions of the target variable are similar across the three datasets
cat("Train set target variable proportions:\n", prop.table(table(datasets$train$label)), "\n\n")
cat("Validation set target variable proportions:\n", prop.table(table(datasets$val$label)), "\n\n")
cat("Test set target variable proportions:\n", prop.table(table(datasets$test$label)), "\n")

# save separate datasets
train <- datasets$train
valid <- datasets$val
test <- datasets$test

# counts
table(train$label)
table(valid$label)
table(test$label)


# save to file - access by other sections independently
write.csv(train, "../data/chem_train.csv", row.names = FALSE)
write.csv(valid, "../data/chem_valid.csv", row.names = FALSE)
write.csv(test, "../data/chem_test.csv", row.names = FALSE)

```

<!--# The above output needs to be inspected to ensure that each subset of the original dataframe is the correct size (proportion) and has the same proportions of target labels, if stratification is required. -->
