---
title: "Chemical Analysis Report"
author: "Petter Lövehagen"
date: "2023-05-07"

format: 
  docx:
    reference_docx: "custom-reference-doc.docx"
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-line-numbers: true
    theme: spacelab
    html-math-method: mathjax
    highlight: "tango"
    #css: "../data/custom.css"
output:
  quarto::word_document:
    reference_docx: "custom-reference-doc.docx"
  html:
    toc: true
    toc-depth: 2
    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-line-numbers: true
    theme: spacelab
    html-math-method: mathjax
    highlight: "tango"
    # #css: "../data/custom.css"
 
lang: "en-GB"
bibliography: "../data/AS_CW_References.bib"
csl: "../data/harvard-university-of-the-west-of-england.csl"
knitr:
  opts_chunk:
    include: false
    tidy: true
    echo: false
    warning: false
    message: false
    error: true
editor: 
  markdown: 
    wrap: 72
---

```{r}
source("packages.R")

```

```{r inspect_data}
# set a seed for reproducibility
set.seed(567)

# load csv file
chem_data <- read.csv("../data/5976423.csv")

# display first rows
head(chem_data)

# check dimensions of df
dim(chem_data)
```

## Data {#sec-data}

The supplied dataset contains `r nrow(chem_data)` measurements of
`r ncol(chem_data)-1` numerical variables.

The analysis team and report writer were not supplied with any
additional information about the data, e.g.:

-   what has been measured
-   variable importance, interactions
-   measurement scale, units, reliability

Therefore, variables are treated equally and without bias in the
analysis and modelling. Decisions are made on statistical grounds and
stated assumptions.

::: callout-tip
## Tip

Additional contextual information about the data could result in
alternative modelling decisions and outcomes and should be considered.
:::

### Missing data {#sec-missing-data}

```{r}
# check for missing data

sum(is.na(chem_data))
mean(is.na(chem_data)) * 100
```

```{r missing_summaries}
# create a new data frame with only rows with missing data
missing_rows <- chem_data[apply(chem_data, 1, function(row) any(is.na(row))), ]

# print the missing rows data frame
print(missing_rows)

# dimensions of missing rows
#dim(missing_rows)

# summarise missing values per column 
na_counts <- colSums(is.na(missing_rows))
  
# calculate percentage of missing values per variable
na_percent <- round(na_counts/nrow(chem_data)*100, 2)

# table
na_var2 <- gt(data.frame(Variable = names(na_counts), Count = na_counts, Percentage = na_percent)) |>
    tab_header(title = "Missing values per variable")

na_var <- gt(data.frame(Variable = names(na_counts), Count = na_counts, Percentage = na_percent)) 



# summarise missing values by label
na_summary <- chem_data %>%
  group_by(label) %>%
  summarize_all(~ sum(is.na(.)))

total_obs <- nrow(chem_data)

row_sums <- data.frame(Label = na_summary$label, 
                        Count = rowSums(na_summary[,-1]),
                        Percentage = round(rowSums(na_summary[,-1])/total_obs * 100, 2))

na_lab2 <- gt(row_sums) |>
  tab_header(title = "Missing values by 'label'") 

na_lab <- gt(row_sums)

na_var2
na_lab2

```

```{r missing_naniar}

# visualise missing pattern
vis_miss(chem_data)

# visualise missing pattern
gg_miss_var(chem_data)

# visualise with facet
gg_miss_fct(chem_data, fct = label)

# pairwise correlation between missingness
gg_miss_upset(chem_data)

# amount and percentage missing for each variable
miss_var<-miss_var_summary(chem_data)

# amount and percentage missing for each observation
miss_case<-miss_case_summary(chem_data)

# visualise with facet
p1 <- gg_miss_fct(chem_data, fct = label)

# pairwise correlation between missingness
p2 <-gg_miss_upset(chem_data)

# extract var with most missing
most_missing_var <- miss_var %>%
  arrange(desc(n_miss)) %>%
  slice(1) %>%
  pull(variable)
  

```

Before proceeding with [Exploratory Data
Analysis](#sec-exploratory-data-analysis), it is important to explore
the extent of missing data and whether there are patterns to the
`missingness`. Some classification models assume no missing data, and
depending on the amount, prevalence and patterning of missing data,
different assumptions and techniques can be applied.

We investigate `missingness` to see whether the assumption that missing
data is 'missing at random' (MAR) holds - that is, is the probability of
`missingness` only dependent on **observed** variables? If so and there
is a significant amount of missing data (\>5%), multiple imputation
approaches may be explored to *impute* missing values by estimation
through statistical inference.

There are `r sum(is.na(chem_data))` rows with at least one missing value
which is `r round(mean(is.na(chem_data)) * 100,2)`% of the entire
dataset. There are significantly fewer missing values than the 5%
threshold, so imputation is not required.

```{r echo=FALSE, include=TRUE,message=FALSE, warning=FALSE, fig.id="fig_visMiss", fig.cap="Figure 1: Missing Data" }
# visualise missing pattern
vis_miss(chem_data)
```

If the missing data is not MAR - that is, it appears to depend on
**unobserved** variables, there may be other, more appropriate
imputation methods such as maximum likelihood imputation. Unobserved
influences in this case may include the sample purity/quality, sample
handling/measuring differences, measurement/equipment discrepancies or
chemical compositions of the samples.

*Imputation* can introduce bias into the analysis if the assumptions of
the imputation method are violated, or the imputed values differ
significantly from true missing values.

::: callout-caution
The client should consider `missingness` in their data:

-   Are there any concerns regarding missing data?
-   Is the equipment, staff, technique, facilities, chemicals, etc.
    consistent?
-   What are the reasons for missing data? Can they be mitigated?
-   Can data be collected where it has been identified as missing?
-   Is there likely to be (more, less, similar amount of) missing data
    in the future?
:::

On a case basis, no single observation has more than
`r max(miss_case$n_miss)` missing value. This means that, at most, a
sample is missing `r round(max(miss_case$pct_miss), 2)`% of its data. If
we group by 'label' we see that `Class E` has the most missing data at
0.52%.

```{r echo=FALSE, include=TRUE,message=FALSE, warning=FALSE, fig.id="fig_missLab", fig.cap="Table 1: Missing Data by 'label'" }
na_lab

```

Variable `r most_missing_var` has the most missing values with
`r miss_var[which.max(miss_var$n_miss), "n_miss"]`, accounting for
`r max(miss_var$pct_miss)`% of this variable's data.

```{r echo=FALSE, include=TRUE,message=FALSE, warning=FALSE, fig.id="fig_missVar", fig.cap="Table 2: Missing Data by Variable" }
na_var

```

The upset plot below shows the five variables with the most missing
values and confirms that there are no instances where there are missing
values in two variables in the same observations.

```{r include=TRUE, echo=FALSE, message = FALSE, fig.id=TRUE, fig.cap="Figure 2: Upset plot of missing data", label="fig_upset", plotid = "fig_upset"}
# call upset plot
p2
```

The target variable ('label') has no missing values and when we group
data by the classes,\
patterns of `missingness` begin to emerge. It appears that missing
values for sequential variables (measurements) may be correlated with
the chemical class. There is a clear stepped pattern in the heatmap
below and the 'Missing Data' plot above.

::: callout-warning
Is there anything systemic, in terms of `missingness`:

-   X1-X4 is *only* missing for chemical A
-   X6, X8 are *only* missing for chemical B
-   X9-X12 is *only* missing for chemical C
-   X13-X16 are *only* missing for chemical D
-   X17-X19 are *only* missing for chemical E

Is there a relationship between the variables, in terms of what they
measure?
:::

```{r include=TRUE, echo=FALSE, message = FALSE, fig.cap="Figure 3: Missing data per variable, grouped by Chemical Class (label)", fig.id=TRUE, label="fig_heatLab"}

p1
```

\[include citation to missing data references\]

::: callout-important
## Decision

Given the minimal amount of missing data - it will be removed before
exploratory data analysis, data splitting and modelling.

If this dataset is representative, the missing data may not be an
issue - assuming the pattern identified above is understood.
:::

### Splitting data

```{r}
clean_chem_data <- chem_data[complete.cases(chem_data), ]
#clean_chem_data
dim(clean_chem_data)
sum(is.na(clean_chem_data))
```

Before proceeding with additional exploratory analysis, we split the
dataset into three random subsets:

-   `train` - used to *train* the classification model to *learn* the
    relationship between the variables and the label
-   `validation` - used to *validate* the performance of the trained
    model and to tune any hyperparameters; determines how well the model
    generalises to 'unseen' data
-   `test` - used to *evaluate* the final performance of the model post
    training and tuning; this data is ony used once and kept separate.

As EDA will explore and visualise relationships between variables,
potentially creating new features, it is essential to split the data so
that we do not introduce bias into the process. Any insights or
observations during the EDA phase will therefore emerge solely from the
training dataset.

The only thing we will check in advance is the target variable balance.
Given that it is not perfectly balanced, the data subsets will be split
with 'stratification' which ensures each partition has a representative
proportion of each class.

```{r include=TRUE, echo=FALSE, message = FALSE, fig.cap="Table 3: Frequency Table", fig.id=TRUE, label="fig_freq"}
#table(clean_chem_data$label)


# frequency table
clean_chem_data %>%
  dplyr::count(label, name = 'Freq') %>% 
  gt() 




```

::: callout-important
## Decision

The clean dataset is sufficiently large to split into:

-   `train` - 50%
-   `validation` - 25%
-   `test` - 25%
:::

```{r split_test_code}
train_prop <- 0.5
val_prop <- 0.25
test_prop <- 0.25


N <- nrow(clean_chem_data)
train_size <- round(N * train_prop)
test_size <- round(N * test_prop)
val_size <- N - train_size - test_size

train_size
test_size
val_size

train_size+test_size+val_size
dim(clean_chem_data)

#unique(clean_chem_data$label)
table(clean_chem_data$label)
```
```{r}
# reusable function to split into three - train, validate, test - finally!!
# this took so long.

df_splitter <- function(df, target_col, train_prop, val_prop, test_prop, stratify) {
  
  if (stratify) {
   
    # split the dataframe into groups based on the target variable
    groups <- split(df, df[[target_col]])
    
    # for each group, randomly sample a portion of the rows for the train set
    train_list <- lapply(groups, function(x) x[sample(nrow(x), floor(train_prop * nrow(x))), ])
    
    # combine the randomly sampled groups into one train set
    train_df <- do.call(rbind, train_list)
    
    # create a dataframe of the remaining rows not in the train set
    val_test_df <- df[!row.names(df) %in% row.names(train_df), ]
    
    # split the remaining rows into groups based on the target variable
    groups_val_test <- split(val_test_df, val_test_df[[target_col]])
    
    # for each group, randomly sample a portion of the rows for the validation set
    val_list <- lapply(groups_val_test, function(x) {
      n <- nrow(x)
      val_indices <- sample(n, floor(val_prop * n), replace = FALSE)
      x[val_indices, ]
    })
    
    # combine the randomly sampled groups into one validation set
    val_df <- do.call(rbind, val_list)
    
    # for each group, randomly sample a portion of the rows for the test set
    test_list <- lapply(groups_val_test, function(x) {
      n <- nrow(x)
      test_indices <- sample(n, floor(test_prop * n), replace = FALSE)
      x[test_indices, ]
    })
    
    # combine the randomly sampled groups into one test set
    test_df <- do.call(rbind, test_list)
    
  } else {
    # Randomly sample indices for each set
    # randomly sample a portion of the rows for the train set
    train_indices <- sample(nrow(df), floor(train_prop * nrow(df)), replace = FALSE)
    train_df <- df[train_indices, ]
    
    # create a dataframe of the remaining rows not in the train set
    val_test_df <- df[!row.names(df) %in% row.names(train_df), ]
    
    # randomly sample a portion of the rows for the validation set
    val_indices <- sample(nrow(val_test_df), floor(val_prop * nrow(val_test_df)), replace = FALSE)
    val_df <- val_test_df[val_indices, ]
    
    # randomly sample a portion of the rows for the test set
    test_indices <- sample(nrow(val_test_df), floor(test_prop * nrow(val_test_df)), replace = FALSE)
    test_df <- val_test_df[test_indices, ]
  }
  
  # Return a list of the three data frames
  return(list(train = train_df, val = val_df, test = test_df))
}






```

```{r}
# apply to clean_chem_data
datasets <- df_splitter(df = clean_chem_data, target_col = "label", train_prop = 0.5, val_prop = 0.25, test_prop = 0.25, stratify = TRUE)

# print the sizes of each resulting dataset
cat("train:", nrow(datasets$train), "\n\n")
cat("validation:", nrow(datasets$val), "\n\n")
cat("test:", nrow(datasets$test), "\n\n\n")

# check that the proportions of the target variable are similar across the three datasets
cat("Train set target variable proportions:\n", prop.table(table(datasets$train$label)), "\n\n")
cat("Validation set target variable proportions:\n", prop.table(table(datasets$val$label)), "\n\n")
cat("Test set target variable proportions:\n", prop.table(table(datasets$test$label)), "\n")

# save separate datasets
train <- datasets$train
val <- datasets$val
test <- datasets$test

# counts
table(train$label)
table(test$label)
table(val$label)



```





### Exploratory Data Analysis {#sec-exploratory-data-analysis}

```{r}
# structure and summary

str(train)
summary(train)




```

```{r}
# overall min
min(train[1:20])
minCol_index <- which(train[1:20] == min(train[1:20]), arr.ind = TRUE)[2]
colnames(train)[minCol_index]

# overall max
max(train[1:20])
maxCol_index <- which(train[1:20] == max(train[1:20]), arr.ind = TRUE)[2]
colnames(train)[maxCol_index]



# ranges
ranges <- apply(train[,-21], 2, function(x) diff(range(x)))
ranges
maxRange <- max(ranges)
minRange <- min(ranges)
maxRange_var <- colnames(train)[which.max(ranges)]
minRange_var <- colnames(train)[which.min(ranges)]


# standard deviations
sd <- apply(train[,-21], 2, sd)
sd
maxSD <- max(sd)
minSD <- min(sd)
maxSD_var <- colnames(train)[which.max(sd)]
minSD_var <- colnames(train)[which.min(sd)]

# means
means <- apply(train[,-21], 2, mean)
means
maxMean <- max(means)
minMean <- min(means)
maxMean_var <- colnames(train)[which.max(means)]
minMean_var <- colnames(train)[which.min(means)]

# variances
vars <- apply(train[,-21], 2, var)
vars
maxVar <- max(vars)
minVar <- min(vars)
maxVar_var <- colnames(train)[which.max(vars)]
minVar_var <- colnames(train)[which.min(vars)]

```

Throughout EDA and model training, we will only use the `train` dataset
which has a total of `r nrow(train)` rows, with a minimum value of
`r min(train[1:20])` (`r colnames(train)[minCol_index]`) and a maximum
value of `r max(train[1:20])` (`r colnames(train)[maxCol_index]`).

The variables have different scales and variances:

| Statistic | Min                             | Max                             |
|------------------------------------|------------------|------------------|
| Mean      | `r minMean` (`r minMean_var`)   | `r maxMean` (`r maxMean_var`)   |
| Variance  | `r minVar` (`r minVar_var`)     | `r maxVar` (`r maxVar_var`)     |
| Range     | `r minRange` (`r minRange_var`) | `r maxRange` (`r maxRange_var`) |

: Table 4: Statistics across Variables

```{r}
library(ggplot2)

ggplot(train, aes(x = X1, fill = label)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ label, nrow = 3)







```

```{r}





# # Freedman-Diaconis rule for binwidth
# 
# iqr <- IQR(train$value)
# n <- nrow(train)
# binwidth <- 2 * iqr / (n^(1/3))
# binwidth
# 
# 
# 
# 
# train %>%
#   select(-label) %>%
#   gather() %>%
#   ggplot(aes(x = value)) +
#   geom_histogram() +
#   facet_wrap(~ key, scales = "free")
```

```{r}
corr_matrix <- cor(train %>% select(-label))

corrplot(corr_matrix, method = "circle")

```

\[histograms\] \[outliers\] \[pair plots\] \[correlation\]
