---
title: "Chemical Sample Classification Report"
subtitle: "Exploratory Data"
author: "Petter Lövehagen"
date: "08 May 2023"

RVersion: 3.6.0 
RStudio: 2023.03.0+386 Cherry Blossom Release
Platform: "x86_64-w64-mingw32/x64 (64-bit)"
OS: "Windows 11 SE"

format: 
  docx: default
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-line-numbers: true
    theme: spacelab
    html-math-method: mathjax
    highlight: "tango"
    #css: "../data/custom.css"
prefer-docx: true
lang: "en-GB"
bibliography: "../data/AS_CW_References.bib"
csl: "../data/harvard-university-of-the-west-of-england.csl"

knitr:
  opts_chunk:
    include: false
    tidy: true
    echo: false
    warning: false
    message: false
    error: true
    fig.align: left
    out.width: '90%'
    fig.width: 6
    fig.height: 4
---



```{r EDA_packages}
# load/install R packages/libraries
# load flextable defaults

file <- "EDA"
source("chem_packages.R")


# what is currently loaded/attached:
#sessionInfo()

# for this section
# library(flextable)
# library(dplyr)
# library(ggplot2)
# library(cowplot)
# library(corrplot)
# library(GGally)

```

```{r load_data}
# load train data set
train <-read.csv("../data/chem_train.csv")



```

```{r train_str}
# structure and summary

head(train)
dim(train)
str(train)
summary(train)
```

### Exploratory Data Analysis {#sec-exploratory-data-analysis}

```{r trainstatistics}
# extract numeric columns
num_train <- train[, -21]

# overall min and max
overall_min <- round(min(num_train),2)
overall_max <- round(max(num_train),2)
minCol_index <- which(num_train == min(num_train), arr.ind = TRUE)[2]
maxCol_index <- which(num_train == max(num_train), arr.ind = TRUE)[2]
minCol_name <- colnames(num_train)[minCol_index]
maxCol_name <- colnames(num_train)[maxCol_index]

# ranges
ranges <- apply(num_train, 2, function(x) diff(range(x)))
maxRange <- round(max(ranges),2)
minRange <- round(min(ranges),2)
maxRange_name <- colnames(num_train)[which.max(ranges)]
minRange_name <- colnames(num_train)[which.min(ranges)]

# standard deviations
sd_vals <- apply(num_train, 2, sd)
maxSD <- round(max(sd_vals),2)
minSD <- round(min(sd_vals),2)
maxSD_name <- colnames(num_train)[which.max(sd_vals)]
minSD_name <- colnames(num_train)[which.min(sd_vals)]

# means
mean_vals <- apply(num_train, 2, mean)
maxMean <- round(max(mean_vals),2)
minMean <- round(min(mean_vals),2)
maxMean_name <- colnames(num_train)[which.max(mean_vals)]
minMean_name <- colnames(num_train)[which.min(mean_vals)]

# variances
var_vals <- apply(num_train, 2, var)
maxVar <- round(max(var_vals),2)
minVar <- round(min(var_vals),2)
maxVar_name <- colnames(num_train)[which.max(var_vals)]
minVar_name <- colnames(num_train)[which.min(var_vals)]


```
Throughout EDA and model training, only the `train` dataset will be analysed.  It has `r nrow(train)` rows, with an overall minimum value of `r overall_min` (`r minCol_name`) and a maximum value of `r overall_max` (`r maxCol_name`). 

The variables have different scales and variances:

<!-- | Statistic | Min                             | Max                             | -->
<!-- |----------------------------------|-------------------|-------------------| -->
<!-- | Mean      | `r minMean` (`r minMean_name`)   | `r maxMean` (`r maxMean_name`)   | -->
<!-- | Variance  | `r minVar` (`r minVar_name`)     | `r maxVar` (`r maxVar_name`)     | -->
<!-- | Range     | `r minRange` (`r minRange_name`) | `r maxRange` (`r maxRange_name`) | -->

<!-- : Table 4: Statistics across Variables -->


```{r minmaxStats, include=TRUE, echo=FALSE}
#| tbl-cap: "Minimum / Maximum Statistics in 'train'"
#| tbl-cap-location: bottom
#| tbl-alt: "Table with minimum and maximum mean, variance and range in train dataset"
#| label: tbl-minmaxStats



# data frame with the statistics and their values
statistics <- data.frame(
  Statistic = c("Mean", "Variance", "Range"),
  "Minimum" = c(paste0(minMean, " (", minMean_name, ") "), paste0(minVar, " (", minVar_name, ") "), paste0(minRange, " (", minRange_name, ") ")),
  "Maximum" = c(paste0(maxMean, " (", maxMean_name, ") "), paste0(maxVar, " (", maxVar_name, ") "), paste0(maxRange, " (", maxRange_name, ") "))
)


statistics |>
  flextable() |> 
  set_caption("Summary Statistics") |>
  autofit() |> 
  bold(j=1)

```
At glance, we can see that there is a significant spread between the variables in terms of mean, variance and range.  In order to mitigate against any one variable having undue influence and because nothing is known about the variables, they will be `normalised` or  standardised so that they have the same scale from 0 to 1.  


```{r}
describe(train)[,c(2:5,8,9)]
```

```{r}
par(mar = c(1, 1, 1, 1))
pairs.panels(train)
```

```{r}



set.seed(567)


# create correlation matrix
corr_matrix <- cor(num_train)

# plot correlation matrix using corrplot
corrplot(corr_matrix)

# plot scatterplot matrix using GGally
#ggpairs(train, progress=FALSE)

```



```{r}
ggplot(train, aes(x = X1, fill = label)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ label, nrow = 3)
```

```{r}
# Create a list of plots, with one plot for each variable
plot_list <- lapply(names(train), function(var) {
  ggplot(train, aes_string(x = var, fill = "label")) +
    geom_density(alpha = 0.5) +
    ggtitle(paste0("Distribution of ", var))
})

# Combine the plots into a grid using the cowplot package
plot_grid(plotlist = plot_list, ncol = 4)

```
```{r}
# order - order of ariables - hierarchical clustering
# method - correlation strength

# Compute the correlation matrix
corr_matrix <- cor(train[, 1:20])

# Plot the correlation matrix
corrplot(corr_matrix, method = "color", type = "upper", order = "hclust",
         col = colorRampPalette(c("#BB4444", "#EE9988", "#FFFFFF", "#77AADD", "#4477AA"))(200),
         addCoef.col = "black", tl.col = "black")
```


Observations
* many normal distributions
* some which differ across labels, eg. X7-X11

```{r}


train %>%
  select(label, X1, X2, X3, X4, X5) %>% 
  GGally::ggpairs(aes(color = label),
          columns = c("X1", "X2", "X3", "X4", "X5"), progress = FALSE) 

```

```{r}

train %>%
  select(label, X7, X8, X9, X10, X11) %>% 
  GGally::ggpairs(aes(color = label),
          columns = c("X7", "X8", "X9", "X10", "X11"), progress = FALSE) 

```

```{r}
corr_matrix <- cor(train %>% select(-label))

corrplot(corr_matrix, method = "circle")
```
## NOTES BELOW

so there appears to be value in two different approaches
* correlation for dimension reduction
* needs domain knowledge to clarify and ratifiy, explain

* feature reduction - several variables which may be superfluous, measuring similar things or do not add to the model



To do
* normalise and scale
* check for outliers 
* address outliers as appropriate
* correlation plot

To try:

```{r}
# correlation plot

corrplot(cor(num_train))
corrplot(cor(num_train), method = "ellipse", type = "lower")
```
interpretation - say what you see
* sparse - lots of variables which are not correlated with each other
* x17-20, and x7-10 appear to correlate, some neg, some pos





### dimension reduction:

PCA - identify principal components, linear combos of original variabes explaining the most variation in the data, plot to see in a lower dimension space
t-SNE

PCA: 

like a plot with PC1, PC2
the actual 5 classes (shape and label by label)
pca arrows and variable labels

(see penguin example)

```{r}
KMO(train[1:20])
```
OVeral KMO is 0.71 which is above 0.5 indicating that the dataset has sufficient correlation present to warrant PCA/FA.  The variable with the lowest ?? is X15 with 0.46.

some citation here.

parallel analysis [citation]

```{r}
scree(num_train)
```
```{r}
paran(num_train, iterations = 5000)
```
```{r}
# save as separate function and call in code
# Define a function to scale/normalize data
scale_data <- function(data) {
  m <- colMeans(data) # means of columns
  MeanCentred <- t(apply(data, 1, function(x){x-m}))
  
  s <- apply(MeanCentred, 2, function(x){sd(x)*(length(x)-1)/length(x)})
  ScaledData <- t(apply(MeanCentred, 1, function(x){x/s}))
  
  return(ScaledData)
}

# Usage example
num_train_scaled <- scale_data(num_train)

```




```{r}
library(ade4)
library(factoextra)
pca <- dudi.pca(num_train,scannf = F, nf=4)
fviz_pca_biplot(pca)
fviz_pca_biplot(pca, label="var",title="Chemical variables", repel=TRUE)
fviz_screeplot(pca)
get_eigenvalue(pca)
fviz_pca_biplot(pca,repel=T,label="var",axes=c(1,2))
fviz_pca_biplot(pca,repel=T,label="var",axes=c(1,3))
```

```{r}

## write your code here 
fviz_pca_biplot(pca,
                repel=T,
                label="var",
                col.ind = train$label)
fviz_pca_biplot(pca,axes=c(1,3),
                repel=T,
                label="var",
                col.ind = train$label)


fviz_pca_ind(pca,
             geom.ind = "point", # show points only (but not "text")
             col.ind = train$label, # colour by groups
             addEllipses = TRUE, # Concentration ellipses
            legend.title = "Chemicals"
            )

fviz_pca_ind(pca,axes=c(1,3),
             geom.ind = "point", # show points only (but not "text")
             col.ind = train$label, # colour by groups
             addEllipses = TRUE, # Concentration ellipses
            legend.title = "Chemicals"
            )
```

















