---
title: "Working file"
subtitle: "Dimension Reduction"
author: "Petter Lövehagen"
date: "08 May 2023"

format: 
  docx: default
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    toc-location: left
    code-fold: true
    code-line-numbers: true
    theme: spacelab
    html-math-method: mathjax
    highlight: "tango"
    #css: "../data/custom.css"
prefer-docx: true
knitr:
  opts_chunk:
    include: false
    tidy: true
    echo: false
    warning: false
    message: false
    error: true
    fig.align: 'left'
    out.width: '90%'
---

```{r working_packages}

# load/install R packages/libraries
# set seed
# load flextable defaults

file <- ""
source("chem_packages.R")

#"dplyr", 
#"flextable"
#"psych"
#"paran"
#"ade4"
#"factoextra"
#"patchwork"
#"dendextend"
#"ggplot2"
#"cluster"
```

```{r}
# load data from csv
train_scaled_label <- read.csv("../data/train_scaled_label.csv")
train_scaled <- train_scaled_label[, -21]

summary(train_scaled_label)
```

## Principal Components Analysis

Having established that at least some variables are linearly correlated, we will consider `PCA` which is where the 'underlying structure' within a dataset is reduced to fewer 'components' or 'dimensions.'

Suitability for `PCA` can be established statistically with the `KMO` test. The scaled dataset has an overall `KMO` value of `r KMO_result$MSA` which is greater than the established threshold of 0.6 for `PCA` suitability. Individual MSA values are summarised below in @tbl-kmo - we can see that only variables three variables have a measure of sampling adequacy less than 0.6, with `r colnames(num_train_scaled)[which.min(kmo_result$MSAi)]` having a value of `r round(min(kmo_result$MSAi),2).`

<!--# KMO test (Kaiser-Meyer-Olkin) is a measure / diagnostic tool to assess PCA suitabiltiy statistically.  It measures sampling adequacy of the variables - that is, how well do the variables represent the 'underlying' factors. KMO produces a value between 0 and 1 where 0.6 or greater is considered acceptable for PCA or factor analysis; less than 0.5 suggest that the dataset may not be suitable.  -->

```{r KMO}
kmo_result <- KMO(train_scaled)
kmo_result$MSA
#KMO_result$Image
kmo_result$MSAi

round(min(kmo_result$MSAi),2)
colnames(train_scaled)[which.min(kmo_result$MSAi)]

```

```{r kmo_summary, include=TRUE, echo=FALSE, message = FALSE, tab.align='left', fig.alt="Summary of individual measures of sampling adequacy (MSA)", tbl}
#| tbl-cap: "Summary of individual measures of sampling adequacy (MSA)"
#| tbl-cap-location: bottom
#| tbl-alt: "Summary of individual measures of sampling adequacy (MSA)"
#| label: tbl-kmo


# summary object, pipe through steps to add colnames, msa band

kmo_summary <- train_scaled %>% 
  KMO() %>% 
  .$MSAi %>% 
  tibble::enframe() %>% 
  mutate(name = colnames(train_scaled), 
         MSAi_band = cut(value, 
                         breaks = c(0, 0.5, 0.6, 0.7, 0.8, 0.9, Inf), 
                         labels = c("< 0.5", "0.5-0.59", "0.6-0.69", "0.7-0.79", "0.8-0.89", ">= 0.9"))) %>% 
  dplyr::select(name, MSAi_band)

# suitability tibble with matching msa banding
suitability <- tibble(
  MSAi_band = c("< 0.5", "0.5-0.59", "0.6-0.69", "0.7-0.79", "0.8-0.89", ">= 0.9"),
  Suitability = c("Unsuitable", "Marginal", "Mediocre", "Good", "Excellent", "Excellent")
)

# summary join
kmo_summary <- kmo_summary %>% 
  left_join(suitability, by = "MSAi_band") %>% 
  group_by(MSAi_band, Suitability) %>% 
  summarise(Variables = paste(name, collapse = ", ")) 

# reorder by msa
kmo_summary$MSAi_band <- factor(kmo_summary$MSAi_band, levels = c(">= 0.9", "0.8-0.89", "0.7-0.79", "0.6-0.69", "0.5-0.59", "< 0.5"))
kmo_summary <- kmo_summary[order(kmo_summary$MSAi_band, decreasing = FALSE), ]


# flextable
flextable(kmo_summary) |>
  width(j = "Variables", width = 6.5, unit = "cm")



```

<!--# Scree plots allow us to visually determine how many 'dimensions' are in the underlying data.  Usually the cutoff is at the 'elbow' joint or where eigenvalues are greater than 1.  The plot below suggests four plots for PCA (and 3 for Factor Analysis), after which additional dimensions do not significantly account for the variability in the dataset. -->

```{r scree}

```

```{r}

```

<!--# Horn's Parallel Analysis has been applied to the dataset, which estimates the number of 'significant' components through the generation of a large number of simulated data sets - in this case 5000.  The resulting eigenvalues are averaged to get a mean estimate of the expected eigenvalues, which are compared to the observed values in order to determine how many 'significant' components to keep. The results below suggest to retain 4 components - based on the eigenvalue being greater than 1. -->

```{r parallel_analysis}

paran(train_scaled, iterations = 5000)
```

<!--# So we do a PCA on the scaled data using dudi from factoextra. By looking at the screeplot of the PCA, it seems that 3 components is more appropriate.  From the 4th component, the contribution to explaining variability is about 5% meaning that each additional component/dimension accounts for approximately an additional 5% of the variability in the dataset.  This is not a lot - so keeping 3 components seems most reasonable.  That said, three components only explain 40% of the variability in the dataset.   -->

```{r pca_plots}

pca_train_scaled <- dudi.pca(train_scaled,scannf = F, nf=4)

scree <- fviz_screeplot(pca_train_scaled)
eigen <- get_eigenvalue(pca_train_scaled)

pca1 <- fviz_pca_biplot(pca_train_scaled, label="var",var.lab = TRUE, col.ind = "#D95F02", alpha.ind = 0.4, col.var = "grey30", repel=TRUE)  +
  ggtitle("PCA Biplots (dimensions 1x2, 2x3) of scaled dataset")

pca2 <- fviz_pca_biplot(pca_train_scaled,label="var", col.ind = "#1B9E77", alpha.ind = 0.4, col.var = "grey30", repel=TRUE,axes=c(2,3)) +
  ggtitle("")

pca3 <- fviz_pca_biplot(pca_train_scaled, col.ind="x", col.var = "grey30", label="var", repel=TRUE,axes=c(1,2)) +
      scale_color_gradient2(low="white", mid="blue",
      high="red", midpoint=0.6)

pca4 <- fviz_pca_biplot(pca_train_scaled,
                        repel = T,
                        label = "var",
                        col.ind = train$label) + 
  theme(legend.position = "none")

pca5 <- fviz_pca_ind(pca_train_scaled,
                     geom.ind = "point", # show points only (but not "text")
                     col.ind = pca_train_scaled$label, # colour by groups
                     addEllipses = TRUE,
                     ggtheme = theme_minimal() + 
                       theme(legend.position = "bottom", 
                             plot.title = element_text(hjust = 0.5, size = 14)))  +
  ggtitle("PCA Biplots with Ellipses by Labels") + 
  theme(legend.position = "bottom", legend.justification = "center", legend.box = "vertical", legend.title = element_text(size = 10), legend.text = element_text(size = 8))


scree
eigen

# pca1
# pca2
# pca3
# pca4
 pca5



(pca1 + pca2)

#/ (pca4 + pca5)

```

<!--# Looking at the biplots of the three dimensions, we can see the correlations identified earlier.  Correlations are interpreted by the angles - 0 degree angles have no correlation with each other (e.g. X17, X19), 180 degree angles between variables are strongly negatively correlated (e.g. X17/X19 with X18/X20) and variables with 90 degree angles are uncorrelated (e.g.X8/X10 with X17, X19). The length of the vector indicates the importance of the variable in explaining the variance in the data.    -->

<!--# PCA5 has plotted the individual points grouped by colour with added ellipses around the centres of the groups.  We can see the clustering by label although there is a significant amount of overlayering.  -->

<!--#  The high degree of overlap between groups suggests that there may not be enough information to effectively differentiate between the different groups.  That the first three dimensions only explain 40% of the variance and each subsequent dimension only adds 5% to this indicates that PCA is probably not going to be helpful.-->

<!--# Distance plot below - shows 'pairwise' distances between observations - darker colours indicate larger distances, lighter colours indicate smaller distances. Clusters of similar observations are darker blocks.   -->

```{r euclid_distance}
# euclidean distance
dist_euc <-get_dist(train_scaled,method="euclidean")

# heatmap
fviz_dist(dist_euc)

```

```{r clusters}
# euclidean distance
dist_euc <- get_dist(train_scaled, method = "euclidean")

# hierarchical clustering on distance using ward.D2 link method
hc <- hclust(dist_euc, method = "ward.D2")

# cuts dendogram to produce five clusters
res_cluster <- cutree(hc, k = 5)


# plot
cluster_5 <- fviz_cluster(list(data = train_scaled, cluster = res_cluster), 
             palette = c("#00AFBB", "#E7B800", "#FC4E07", "#800080", "#808000")
, 
             ggtheme = theme_minimal(), 
             geom = "point")


# cuts dendogram to produce three clusters
res_cluster <- cutree(hc, k = 3)


# plot
cluster_3 <- fviz_cluster(list(data = num_train_scaled, cluster = res_cluster), 
             palette = c("#00AFBB", "#E7B800", "#FC4E07")
, 
             ggtheme = theme_minimal(), 
             geom = "point")

cluster_3
cluster_5
```

<!--# Clustering plot based on Euclidean distance calculated from the dataset.  Scatterplot of dimensions (1, 2) and a hierarchical clustering algorithm applied. Resulting dendograms are visualisations where x-axis represents individual observations and y-axis represents the distance between them.  The height where two branches merge is the distance between them. -->

```{r hclust_single}
# nearest neighbour
NN<-hclust(dist_euc,method="single")
#plot(NN)

# labels from the original data
labels <- rownames(train_scaled)

# labels to the dendrogram
NN$labels <- labels

# dendrogram with labels
plot(NN, cex = 0.8, hang = -1)


```

<!--# The above plot is 'single linkage' / nearest neighbour using hclust().  It is a clustering method where distance between two clusters is defined as the shortest distance between any two points in the two clusters.  The unintelligible dendogram shows how the data points are grouped into clusters based on similarity.  The vertical height represents the distance between samples/clusters.  -->

```{r hclust_complete}
# complete linkage
comp_clust <-hclust(dist_euc,method="complete")

# labels to the dendrogram
comp_clust$labels <- labels

plot(comp_clust, cex = 0.8, hang = -1)
```

<!--# The resulting dendogram from 'complete linkage' method shows more of a tree structure to the clustering.  Complete linkage measures distance between clusters by considering the maximum distance between any two points in the clusters.  The top level splits are easier to visualise. -->

```{r hclust_average}
# average link method
AD<-hclust(dist_euc,method="average")

plot(AD)
```

```{r hclust_wards}
# wards method
WC<-hclust(dist_euc,method="ward.D2")
WC$labels <- labels

plot(WC)
```

<!--# Clustering with Wards method (minimum variance method with squared Euclidean distance) progressively creates groups based on similarity.  The method tries to minimise sum of squared differences within each cluster - so compact clusters become identifiable.  The above dendrogram is much more useful than earlier dendrograms - there are clear 3-5 groups. -->

```{r Kmeans}
# k-means
# run k-means with 3 and 4 clusters
#KM3 <- kmeans(train_scaled, 3)
#KM4 <- kmeans(train_scaled, 4)
KM5 <- kmeans(nrain_scaled, 5)

# create a colour palette with the same number of colors as clusters
library(RColorBrewer)
col3 <- brewer.pal(3, "Set1")
col4 <- brewer.pal(4, "Set1")
col5 <- brewer.pal(5, "Set1")

# create scatter plots with colors and symbols based on cluster membership
par(mfrow = c(1, 3))
plot(train_scaled[, 1:2], col = col3[KM3$cluster], pch = KM3$cluster,
     xlab = "", ylab = "PC2", main = "k-means (k = 3)")
plot(train_scaled[, 1:2], col = col4[KM4$cluster], pch = KM4$cluster,
     xlab = "PC1", ylab = "", main = "k-means (k = 4)")
plot(train_scaled[, 1:2], col = col5[KM5$cluster], pch = KM5$cluster,
     xlab = "", ylab = "", main = "k-means (k = 5)")

```

<!--# K-means clustering is another clustering method which selects initial 'cluster centroids' for the number of specified clusters.  The algorithm iteratively recalculates the means as points are added. It is difficult to see with any clarity the clusters, although there are groupings where they converge.-->

/

```{r tanglegram}

library(dendextend)
library(ggplot2)

# Create the tanglegram plot
my_tanglegram <- tanglegram(comp_clust, NN)

# Save the plot as a PNG file
ggsave("tanglegram.png", my_tanglegram, width = 8, height = 6, dpi = 300)

```

<!--# Like parallel analysis, nbclust is a function to determine how many clusters are in the dataset.  Parallel analysis calculates and plots several clustering validation indices (Calinski-Harabasz, silhouette width and Davies-Bouldin) for differen values of K, summarising and choosing the optimal number of clusters.  Nbclust calculates and plots the `within-cluster sum of squares (WSS) for different values of K (clusters) and uses the elbow method.  The plot below suggests 3 or 4 clusters.-->

```{r wss_clust}
# within cluster sum of squares
fviz_nbclust(train_scaled,FUN=hcut,method="wss")
```

<!--# The 'silhouette' method calculates the average silhouette width for different values of k (clusters) and plots it.  The silhouette width measures similarity to a cluster in comparison to other clusters.  A high average silhouettte indicates that data is well-clustered; a low value suggests that there is overlap or misclassification.  The plot below suggests 2 clusters. -->

```{r silhouette_clust}
fviz_nbclust(train_scaled, FUN = hcut, method = "silhouette")
```

```{r km_wss}
# how far points within cluster are from centre of cluster - should be minimised
KM5$withinss
```

```{r km_tot_wss}
# sum of wss
KM5$tot.withinss
```

```{r km_betweenss}
# sum of squared distances between cluster centres and overall mean
# high value indicates more separation between clusters
KM3$betweenss
```

```{r km_size}
# size of each cluster
KM5$size
```

```{r cluster_KM}
fviz_cluster(KM4,data = train_scaled, repel = TRUE, geom = "point", label = "none")
fviz_cluster(KM5, data = train_scaled, repel = TRUE, geom = "point", label = "none")

```

```{r}


# Compute the silhouettes for different k values
sil <- sapply(2:10, function(k) {
  km <- kmeans(train_scaled, centers = k)
  avg_sil <- mean(silhouette(km$cluster, dist_euc))
  return(avg_sil)
})

# Plot the silhouette values for each k
plot(2:10, sil, type = "b", xlab = "Number of clusters (k)", ylab = "Average silhouette width")

```

```{r}
library(factoextra)
library(cluster)

# Compute within-cluster sum of squares (wss) for different k values
wss <- sapply(2:10, function(k) {
  km <- kmeans(train_scaled, centers = k)
  return(km$tot.withinss)
})

# Plot the wss values for each k
plot(2:10, wss, type = "b", xlab = "Number of clusters (k)", ylab = "Within-cluster sum of squares (wss)")

# Identify the elbow point using the knee locator function
fviz_nbclust(train_scaled, kmeans, method = "wss") + 
  geom_vline(xintercept = 3, linetype = "dashed")

```

```{r}
library(factoextra)
library(cluster)

# Compute gap statistic for different k values
gap_stat <- clusGap(train_scaled, FUN = kmeans, K.max = 10, B = 50, verbose = FALSE)

# Plot the gap statistic values for each k
plot(gap_stat, main = "Gap statistic", xlab = "Number of clusters (k)")

# Identify the optimal number of clusters using the gap statistic function
fviz_gap_stat(gap_stat)


```

<!--# Everything in the clustering / PCA analysis - which is about understanding the underlying structure and patterns of the data - suggests that there are 3 or maybe 4 clusters / dimensions.  However, there are 5 according to the client.  This suggests that there may be more nuance or complexity than this analysis is showing.    -->

<!--#  PCA has not really identified any variables to drop as they are all contributing a little.  -->

```{r}
library(factoextra)

# PCA
pca <- prcomp(train_scaled)

# loadings for all principal components
loadings <- get_pca_var(pca)
loadings$contrib
loadings$cor

# Get variable names with low contributions
low_contrib <- loadings$contrib[abs(loadings$contrib) < 0.2]
if (length(low_contrib) == 0) {
  cat("No variables with contributions less than 0.2\n")
} else {
  low_loadings <- names(low_contrib)
  cat("Variables with low contributions (<0.2):\n")
  cat(paste(low_loadings, collapse=", "))
}
#low_contrib


```

```{r}




# create the data frame
var_contrib <- data.frame(
  Variable = names(train_scaled),
  Loading = round(loadings$contrib, 2)
)


# create a flextable object
flextable(var_contrib)



```

```{r}




```
