---
title: "Chemical Analysis Report"
subtitle: "Classification Model"

author: "Petter Lövehagen - petter2.lovehagen@live.uwe.ac.uk"
date: "07 May 2023"

RVersion: 3.6.0 
RStudio: 2023.03.0+386 Cherry Blossom Release
Platform: "x86_64-w64-mingw32/x64 (64-bit)"
OS: "Windows 11 SE"



# output:
#   html_document:
#     toc: true
#     toc_depth: 2
#     toc_title: "Report Contents"
#     toc_float: true
#     code_folding: hide
#     theme: spacelab
#     highlight: tango
#     number_sections: true
#     df_print: paged
#   pdf_document:
#     latex_engine: pdflatex
#     number_sections: true
#     toc: true
#     toc_depth: 2
#     toc_title: "Report Contents"
#   word_document:
#     reference_docx: custom-reference-doc.docx
#     keep_md: true
#     keep_tex: true

# output:
#   quarto::quarto_docx:
#     reference_docx: custom-reference-doc.docx
#     template: custom-reference-doc.docx
#     keep_md: true
#     keep_tex: true



format: 
  docx: 
    #toc: true
    #toc-depth: 2
    #toc-title: "Report Contents"
    #number-sections: true
    highlight-style: github
    reference-doc: custom-reference-doc.docx
  html:
    toc: true
    toc-depth: 2
    toc-title: "Report Contents"
    toc-location: left
    code-fold: true
    code-line-numbers: true
    theme: spacelab
    html-math-method: mathjax
    highlight: "tango"
    #css: "../data/custom.css"
  pdf:
    latex_engine: pdflatex
    geometry:
      - top=30mm
      - left=30mm
  #ipynb: default
  
prefer-docx: true
lang: "en-GB"
bibliography: "../data/AS_CW_References.bib"
csl: "../data/harvard-university-of-the-west-of-england.csl"
knitr:
  opts_chunk:
    include: false
    tidy: true
    echo: false
    warning: false
    message: false
    error: true
    #comment: "NA"
---

```{r packages}
# "ggplot2", "ggrepel", "forecast", "patchwork"

# vector of packages to install/load
packages <- c("citr", "readr", "naniar", "gt")

for (package in packages) {
  if (!require(package, character.only = TRUE)) {
    if (package == "citr") {
      devtools::install_github("crsh/citr")
    } else {
      install.packages(package)
    }
    library(package, character.only = TRUE)
  }
}

```



## Brief {#sec-brief}

The brief is to develop a classification tool to classify chemicals.  

In addition, the client is interested in knowing whether all variables are necessary for achieving optimal out-of-sample classification performance, or if a subset of measurements will suffice.

This report will aim to: 

* analyse and feedback on the provided dataset
* develop an optimsed classification model
* determine whether all variables are required

## Data {#sec-data}

```{r data}

# load csv file
chem_data <- read.csv("../data/5976423.csv")

# display first rows
head(chem_data)

# check dimensions of df
dim(chem_data)
```


### Missing data {#sec-missing-data}

### Exploratory Data Analysis {#sec-exploratory-data-analysis}

### Splitting data {#sec-splitting-d}

## Model {#sec-model}

### Features {#sec-features}

### Build {#sec-build}

### Evaluation {#sec-evaluation}

### Validation {#sec-validation}

## Summary {#sec-summary}
