---
title: "Missing Data - Notes"
author: "p lovehagen"
format: html
---

```{r}
chem_data <- read.csv("../data/5976423.csv")
```

[CRAN Task View: Missing Data (r-project.org)](https://cran.r-project.org/web/views/MissingData.html)

EDA

Likelihood based approaches

Single imputation

Multiple imputation

Weighting Methods

```{r naniar_missing}
library(naniar)

# Create a visualization of the missing data pattern
vis_miss(chem_data)

# Check if there is any pairwise correlation between missingness
miss_var_summary(chem_data)

# Compute the amount and percentage of missing data for each variable
miss_var_summary(chem_data)

# Compute the amount and percentage of missing data for each observation
miss_case_summary(chem_data)



```

```{r naniar}
library(naniar)

# Create a visualization of the missing data pattern
gg_miss_var(chem_data)

# Check if there is any pairwise correlation between missingness
#gg_miss_upset(chem_data)

# Compute the amount and percentage of missing data for each variable
miss_var_summary(chem_data)

# Compute the amount and percentage of missing data for each observation
miss_case_summary(chem_data)

```

```{r}
library(naniar)

gg_miss_fct(chem_data, fct = label)

```

```{r}
# Load the naniar package
library(naniar)

# Select the columns containing the numeric features
feature_cols <- paste0("X", 1:20)

# Create a subset of chem_data containing only the feature columns
chem_data_features <- chem_data[, feature_cols]

# Create an upset plot of missing values in the feature columns
gg_miss_upset(chem_data_features)

```

```{r}
# Load ggplot2 package
library(ggplot2)

# Create line chart of X1 vs X2, grouped by label
ggplot(chem_data, aes(x = X1, y = X2, group = label)) +
  geom_line()

```

```{r mi}

library(mi)

# Compute the amount and percentage of missing data for each variable
summary(chem_data)

library(mi)
library(VIM)

# Create a missingness matrix
mm <- marginplot(chem_data, "cex.axis" = 1, "las" = 3)

# Visualize the missing data pattern for each variable
aggr(chem_data, col=c('grey','red'), numbers=TRUE, sortVars=TRUE, labels=names(chem_data), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))



```

```{r wrangle}
library(wrangle)

# Create a missingness summary
miss_summary(chem_data)

```

```{r DescTools}
# Load the DescTools package
library(DescTools)

# Generate a missingness matrix plot
Desc(chem_data, plot = TRUE)


```

```{r}
#install.packages("missMDA")
library(missMDA)

# First, remove the label column
chem_data_no_labels <- chem_data[, -ncol(chem_data)]

# Impute missing values using PCA
imputed_data <- imputePCA(chem_data_no_labels)

# Combine the imputed data with the label column
imputed_data_with_labels <- cbind(imputed_data, labels = chem_data$labels)

# Check if there are any missing values left
sum(is.na(imputed_data_with_labels))

# Note that the "labels" column will still have missing values since we did not use it for imputation


```

```{}
```

1.  Assume MAR - missing at random - then multiple imputation (MI) may be appropriate. It can estimate the missing vlaues based on observed data and provide statistical inferences. MAR is where missingness depends only on the observed variables.
2.  If not MAR - depends on unobserved variables like chemical composition of the samples or the measurement process, then other imputation, like maximum likelihood imputation (MLI) or regression may be more appropriate.
3.  Or not so much missing data - so remove those rows?

Imputation can introduce bias into the analysis if the assumptions underlying the imputation model are violated or the imputed values are substantially different from the true missing values...
