---
title: "week 11 - final week tut notes"
format: html
---

## Requirements:

Look at FlexTable for tables

Look at Missing stats from naniar for table production

e.g. miss_var_summary, miss_var_which

only display the missing rows in the table

remove packages which are not used

No document template

consider reproducibility - so if there are no missing values, make an if statement...

Generalise the first section - the EDA.

No expectation of rotation in FA because this requires domain knowledge (which we don't have in this instance)

compliled file

add alternative text to all figures

TODO:

sort out label: fig-figname; fig-cap: "fig title"; label: tbl-tableName; tbl-cap: "table title"

this should allow for @referencing within document

@tbl-tableName

reference different data subsets - tvt

Feature: FlexTable

-   understand

-   set FT style, and defaults for all tables in the report

-   change current tables to FT format -\> replace GT (save as scrap)

Create repeatable missing checks:

-   missing vars (length of missing vars - case when)

Revisit comments

Create bibliography -\> base R, critical packages

Reference comments for 'additional detail' - if client wants

Consider implementing a DataValidateR option in github

-   upload csv, validate for errors

-   upload csv, run classification model

```{r}
# flextable example

x<-cov(airquality, use = "pairwise.complete.obs")
# will turn all numbers into strings- so consider rounding, etc. first
x1 <- cbind(variables = rownames(x), x)
x1 <- as.data.frame(x1)
library(flextable)
flextable(x1)

x1[,-1] <-x
flextable(x1)|>
  colformat_double(digits=2)|>theme_booktabs(bold_header=TRUE) |>
  bold(j=1)|>
  autofit()
```

Dimension reduction:

-   FA, PCA, Parallel analysis - Review tutorials and plan out approach

-   PCA may not be the best approach - eigenvectors/values (not explaining much)

    means keeping all variaables - projecting into dimensions

-   Trees, forests may be more appropriate -\> feature selection (can remove features)

one approach using whole dataset -\> projection (PCA + classification)

other method is feature reduction (RF, for example)

compared in the second dataset 'test'

```{r}
MySentence1<- case_when(
  length(MissingVars) ==0 ~ "there are no missing variables.",
  length(MissingVars) ==1 ~ paste0("the following variable has missing values: ", MissingVars,"."),
  length(MissingVars) ==2 ~ paste0("the following variables have missing values: ", paste(MissingVars, collapse = "and" )), 
  length(MissingVars) >2 ~ paste0("there are ", length(MissingVars))
)
```

if(!is.null(miss_var_which(airquality)),

paste0("the following variables have missing variables:", miss_var_which(airquality)"."

Reporting missing variables - swap into R chunk with rules

missingVars \<- missing_var_which(df)

case_when(

missingVars == 0 - 'no missing'

missingVars ==1 - 'there is one var with missing;

missingVars ==2-5

missingVars \>5

Final file production - ZIP

Need to zip all dependencies - files

-   sections to knit

-   bibliography/references

-   packages.R
